### 2023-09-13 doc/_bayesian_inference.qmd
finished adding Griffin's comments

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, as depicted in @fig-01(C),
the statistical dependencies of our observations and latent variables allow us
to factorize the joint distribution into the product of multiple conditional
distributions. To exemplify this, let us focus on the variable
$\underline{\underline{F}}^M$ within the diagram in @fig-01(C). The incoming
arrows from $\underline{\bar{s}}_T$ and $\underline{s}^M$

rewrite the joint probability distribution as a product of conditional
distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M,
    \underline{\underline{R}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N,
    \underline{\underline{R}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each term. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. We invite the reader to check the supplementary material, where we
work through these terms to build the posterior distribution to sample
from. We refer the reader to `[ref SI section]` for an extended description of
the model with specific functional forms for each term on the left-hand side of
@eq-split_posterior as well as `[ref SI on hierarchical]` where the model is
extended to account for multiple experimental replicates.

@@

To overcome this computational limitation, in this work, we rely on the recent
development of the automatic differentiation variational inference algorithm
(ADVI) [@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta \mid
\underline{\underline{R}})$, where $\phi$ fully parametrizes the approximate
distribution. As further explained in `[point to SI section]`, the numerical
integration problem is replaced by an optimization problem of the form
$$
q^*_\phi(\theta \mid \underline{\underline{R}}) = \min _\phi
D_{KL}(
    q_\phi(\theta \mid \underline{\underline{R}}) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the 
complicated high-dimensional numerical integration problem is transformed into
a gradient descent over the parameters of the approximate posterior distribution
$\phi$ such that it minimizes the KL divergence with respect to the true 
posterior distribution. Although to compute @eq-vi_objective, we require the
posterior distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the 
variational free energy [@buckley2017]---is mathematically equivalent to 
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to `[point to SI section]` for a short primer on variational inference.

>>>>

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, the statistical dependencies
of our observations and latent variables allow us to factorize the joint
distribution into the product of multiple conditional distributions. To gain
some intuition about this factorization, let us focus on the inference of the
population mean fitness values $\underline{\bar{s}}_T$. @eq-logfreq_neutral
relates the value of the population mean fitness to the neutral lineage
frequencies and nothing else. This suggests that when writing the posterior for
these population mean fitness parameters, we should be able to condition it only
on the neutral lineage frequency values, i.e., $\pi(\underline{\bar{s}}_T \mid
\underline{\underline{F}}^N)$. We point the reader to `[SI section]` for the
full mathematical details on this factorization. For our purpose here, it
suffices to say we can rewrite the joint probability distribution as a product
of conditional distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each term. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. We invite the reader to check the supplementary material, where we
work through these terms to build the posterior distribution to sample
from. We refer the reader to `[ref SI section]` for an extended description of
the model with specific functional forms for each term on the left-hand side of
@eq-split_posterior as well as `[ref SI on hierarchical]` where the model is
extended to account for multiple experimental replicates.

@@

To overcome this computational limitation, in this work, we rely on the recent
development of the automatic differentiation variational inference algorithm
(ADVI) [@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta)$, where $\phi$
fully parametrizes the approximate distribution. As further explained in
@sec-vi_primer, the numerical integration problem is replaced by an optimization
problem of the form
$$
q^*_\phi(\theta) = \min _\phi
D_{KL}(
    q_\phi(\theta) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the
complicated high-dimensional numerical integration problem is transformed into a
much simpler problem of finding the value of the parameters $\phi$ such that
@eq-vi_objective is satisfied as best as possible within some finite computation
time. Although to compute @eq-vi_objective, we require the posterior
distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the
variational free energy [@buckley2017]---is mathematically equivalent to
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to @sec-vi_primer for a short primer on variational inference.

### 2023-09-13 doc/_experiment.qmd
finished adding Griffin's comments

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time to then
dilute the culture back into fresh media and repeat for roughly 4-7 cycles,
depending on the initial abundances of the mutants, and their relative growth
rates. The DNA barcodes are sequenced at the end of each growth cycle to
quantify the relative abundance of each of the barcodes. We point the reader to
@kinsler2020 for specific details on these assays for *S. cerevisiae* and to
@ascensao2023 for equivalent assays for *E. coli*. @fig-01(B) presents a typical
barcode trajectory where the darker barcode trajectories represent the so-called
*neutral lineages* that are genetically equivalent to the untagged ancestor
strain that initially dominates the culture. These neutral lineages simplify the
inference problem since the fitness of all relevant barcodes is quantified with
respect to these barcodes.

![**Typical competitive fitness experiment**. (A) Schematic of the typical
experimental design to determine the competitive fitness of an ensemble of
barcoded genotypes. Genotypes are pooled together and grown over multiple
growth-dilution cycles. At the end of each cycle, a sample is processed to
generate a library for amplicon sequencing. (B) Typical barcode trajectory
dataset. From each time point, the relative frequency of each barcode is
determined from the total number of reads. Different shades of blue represent
different relative fitness. Darker gray lines define the typical trajectory of
neutral lineages. (C) Bayesian network defining the statistical dependence
between observations (filled circles) and latent variables (empty circles). The
dotted circle highlights the set of variables we are interested in. One-way
arrows depict the statistical dependence between variables. For example, the
value of $\underline{\underline{F}}^N$ only depends on the value of
$\underline{\bar{s}}_T$, while $\underline{\underline{F}}^M$ depends on
$\underline{\bar{s}}_T$ and $\underline{s}^M$. See @sec-bayesian_inference for
specific definitions of each variable.](./figs/fig01){#fig-01}

>>>>

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time; then, an
aliquot is inoculated into fresh media for the next growth cycle. This process
is repeated for roughly 4-7 cycles, depending on the initial abundances of the
mutants and their relative growth rates. The DNA barcodes are sequenced at the
end of each growth cycle to quantify the relative abundance of each of the
barcodes. We point the reader to @kinsler2020 for specific details on these
assays for *S. cerevisiae* and to @ascensao2023 for equivalent assays for *E.
coli*. @fig-01(B) presents a typical barcode trajectory where the darker barcode
trajectories represent the so-called *neutral lineages* that are genetically
equivalent to the untagged ancestor strain that initially dominates the culture.
These neutral lineages simplify the inference problem since the fitness of all
relevant barcodes is quantified with respect to these barcodes.

![**Typical competitive fitness experiment**. (A) Schematic of the typical
experimental design to determine the competitive fitness of an ensemble of
barcoded genotypes. Genotypes are pooled together and grown over multiple
growth-dilution cycles. At the end of each cycle, a sample is processed to
generate a library for amplicon sequencing. (B) Typical barcode trajectory
dataset. From each time point, the relative frequency of each barcode is
determined from the total number of reads. Different shades of blue represent
different relative fitness. Darker gray lines define the typical trajectory of
neutral lineages.](./figs/fig01){#fig-01}

### 2023-09-13 doc/_fitness_model.qmd
finished adding Griffin's comments

>>>>

The presence of the neutral lineages facilitates the determination of the 
population mean fitness value $\bar{s}_t$. Since every relative fitness is 
determined relative to the neutral lineage that dominates the culture, we define
their fitness to be $s^{(n)} = 0$, where the subscript $(n)$ specifies their
neutrality. This means that @eq-logfreq for a neutral lineage takes the simpler
form
$$
\frac{1}{\tau}\ln \frac{f_{t+1}^{(n)}}{f_{t}^{(n)}} = - \bar{s}_t.
$${#eq-logfreq_neutral}
Therefore, we can use the data from these reference barcodes to directly infer
the value of the population mean fitness.

### 2023-09-13 doc/_single_results.qmd
finished adding Griffin's comments

To assess the inference pipeline performance we applied it to a simulated
dataset with known ground truth relative fitness values (See `[ref to SI]` for
details on simulation). @fig-02(A) shows the structure of the synthetic dataset.
The majority of barcodes of interest (dim color lines) are adaptive compared to
the neutral barcodes ($s^{(m)} > 0$). This experimental condition is required to
reliably measure the relative fitness since all barcodes of interest occupy a
small fraction of the population at the beginning of the experiment `[ref SI
section on detrimental mutants]`. Although the barcode frequency trajectories
look relatively smooth, our fitness model requires the computation of the log
frequency ratio between adjacent time points as derived in @eq-logfreq.
@fig-02(B) shows such data transformation where we can better appreciate the
observational noise input into our statistical model. This noise is
evident for the darker lines representing the neutral barcodes since
all of these lineages are assumed to be identically distributed under this log
frequency ratio transformation.

A way to visualize the performance of our inference pipeline in fitting our
fitness model to the observed data is to compute the so-called posterior
predictive checks (PPC). In short, the PPC consists of bootstrapping the
generation of synthetic datasets in agreement with the results from the
inference results. In other words, we use the resulting parameter values from
the ADVI inference to generate possible datasets in agreement with the inferred
values. @fig-02(C) shows these results for all neutral lineages (upper left
corner plot) and a few representative non-neutral barcodes. The dark color line
shows the median value we would expect the data to be found, while the shaded
regions show the 68\% and 95\% credible regions, i.e., the regions where we
expect to find the data with the corresponding probability---or in terms of our
parameter, the $X\%$ credible region is the interval where we expect the true
parameter to lie with $X\%$ probability. The main advantage of our method is
this natural interpretability of these credible regions since a common mistake
in the literature is to interpret frequentist confidence intervals as Bayesian
credible regions when they are not equivalent [@morey2016].

To capture the global performance of the model, @fig-02(D) compares the known
ground truth with the inferred relative fitness value for all barcodes of
interest. There is an excellent degree of correspondence between these values,
with the error bars, representing the 68\% credible region for the parameter
value crossing the identity line for most barcodes. This latter point is made
clear with @fig-02(E) where $\approx 90\%$ of ground truth fitness values fall
within one standard deviation of the mean in the inferred posterior
distributions.

![**Single dataset inference**. (A) Frequency trajectories that represent the
raw data going into the inference. (B) Log frequency ratio between two adjacent
time points used by the inference pipeline. Darker lines represent the neutral
barcodes. Notice that these transformed data are much more noisy than the
seemingly smooth frequency trajectories. (B) Examples of the posterior
predictive checks for all neutral lineages (upper left panel) and a subset of
representative mutant lineages. Error bars represent the 68\% credible region on
the mutant relative fitness marginal distribution. (C) Comparison between the
ground truth fitness value from the logistic-growth simulation and the inferred
fitness value. Gray error bars represent the 68\% posterior credible region. (D)
The empirical cumulative distribution function (ECDF) for the absolute z-score
value of the ground truth parameter value within the inferred fitness
posterior distribution.](./figs/fig02){#fig-02}

>>>>

To assess the inference pipeline performance, we applied it to a simulated
dataset with known ground truth relative fitness values (See `[ref to SI]` for
details on simulation). @fig-02(A) shows the structure of the synthetic dataset.
The majority of barcodes of interest (faint color lines) are adaptive compared
to the neutral barcodes ($s^{(m)} > 0$). This experimental condition is required
to reliably measure the relative fitness since all barcodes of interest occupy a
small fraction of the population at the beginning of the experiment `[ref SI
section on detrimental mutants]`. Although the barcode frequency trajectories
look relatively smooth, our fitness model requires the computation of the log
frequency ratio between adjacent time points as derived in @eq-logfreq.
@fig-02(B) shows such data transformation where we can better appreciate the
observational noise input into our statistical model. This noise is evident for
the darker lines representing the neutral barcodes since all of these lineages
are assumed to be identically distributed under this log frequency ratio
transformation.

To visualize the performance of our inference pipeline in fitting our fitness
model to the observed data, we compute the so-called posterior predictive checks
(PPC). In short, the PPC consists of repeatedly generating synthetic datasets in
agreement with the results from the inference results. In other words, we use
the resulting parameter values from the ADVI inference to generate possible
datasets in agreement with the inferred values. @fig-02(C) shows these results
for all neutral lineages (upper left corner plot) and a few representative
non-neutral barcodes. The different color shades represent the 95\%, 68\%, and
5\% credible regions, i.e., the regions where we expect to find the data with
the corresponding probability---or in terms of our parameter, the $X\%$ credible
region is the interval where we expect the true parameter to lie with $X\%$
probability.

The main advantage of our method is this natural interpretability of these
credible regions where an $X\%$ credible region indeed captures the region of
parameter space where we expect with $X\%$ probability the actual value of the
parameter lies given our statistical model, our prior information, and the
observed experimental data. A common mistake in the literature is interpreting
frequentist confidence intervals as Bayesian credible regions when they are not
equivalent [@morey2016]. Frequentist confidence intervals and Bayesian credible
regions are based on fundamentally different philosophical approaches to
statistics. Frequentist confidence intervals represent the range of values that
would contain the true population parameter with a certain probability if the
experiment was repeated many times. The confidence level does not represent the
probability that the interval contains the true value. According to a specific
model and prior information, Bayesian credible regions represent the range of
values that contain the parameter with a certain posterior probability. The
credible region directly represents the probability that the region contains the
true value. So, frequentist confidence intervals cannot be interpreted as
Bayesian credible regions because they have fundamentally different meanings.
Treating an $X\%$ confidence interval like an $X\%$  credible region is
fallacious since confidence levels do not represent probabilistic coverage of
the true value like credible regions. The intervals are generated through
entirely different procedures.

To capture the global performance of the model, @fig-02(D) compares the known
ground truth with the inferred relative fitness value for all barcodes of
interest. There is an excellent degree of correspondence between these values,
with the error bars representing the 68\% credible region for the parameter
value crossing the identity line for most barcodes. This latter point is made
clear with @fig-02(E) where $\approx 90\%$ of ground truth fitness values fall
within one standard deviation of the mean in the inferred posterior
distributions.

![**Single dataset inference**. (A) Frequency trajectories that represent the
raw data going into the inference. (B) Log frequency ratio between two adjacent
time points used by the inference pipeline. Darker lines represent the neutral
barcodes. These transformed data are much more noisy than the
seemingly smooth frequency trajectories. (B) Examples of the posterior
predictive checks for all neutral lineages (upper left panel) and a subset of
representative mutant lineages. Error bars represent the 68\% credible region on
the mutant relative fitness marginal distribution. (C) Comparison between the
ground truth fitness value from the logistic-growth simulation and the inferred
fitness value. Gray error bars represent the 68\% posterior credible region. (D)
The empirical cumulative distribution function (ECDF) for the absolute z-score
value of the ground truth parameter value within the inferred fitness
posterior distribution.](./figs/fig02){#fig-02}

### 2023-09-13 doc/_variational_inference_primer.qmd
finished adding Griffin's comments

## Primer on Variational Inference

@@

To maximize the right-hand side of @eq-vi_elbo, the Automatic Differentiation
Variational Inference (ADVI) algorithm developed in [@kucukelbir2016] takes
advantage of advances in probabilistic programming languages to generate a
robust method to perform this optimization. Without going into the details of
the algorithm implementation, for our purposes, it suffices to say that we define
our joint distribution $\pi(\underline{\theta}, \underline{x})$ as the product
defined in @eq-vi_joint_dist. ADVI then proposes an approximate variational
distribution $q_\phi$ that can either be a multivariate Normal distribution with
a diagonal covariance matrix, i.e.,
$$
\phi = (\underline{\mu}, \underline{\underline{I}}\;\underline{\sigma}^2),
$${#eq-vi_meanfield}
where $\underline{\underline{I}}$ is the identity matrix, and
$\underline{\sigma}$ is the vector of standard deviations or a full-rank 
multivariate Normal distribution
$$
\phi = (\underline{\mu}, \underline{\underline{\Sigma}}).
$${#eq-vi_full}

>>>>

## Primer on Variational Inference{#sec-vi_primer}

@@

To maximize the right-hand side of @eq-vi_elbo, the Automatic Differentiation
Variational Inference (ADVI) algorithm developed in [@kucukelbir2016] takes
advantage of advances in probabilistic programming languages to generate a
robust method to perform this optimization. Without going into the details of
the algorithm implementation, for our purposes, it suffices to say that we define
our joint distribution $\pi(\underline{\theta}, \underline{x})$ as the product
defined in @eq-vi_joint_dist. ADVI then proposes an approximate variational
distribution $q_\phi$ that can either be a multivariate Normal distribution with
a diagonal covariance matrix, i.e.,
$$
\phi = (\underline{\mu}, \underline{\underline{D}}),
$${#eq-vi_meanfield}
where $\underline{\underline{D}}$ is the identity matrix, with the diagonal
elements given by the vector of variances $\underline{\sigma}^2$ for each
variable or a full-rank multivariate Normal distribution
$$
\phi = (\underline{\mu}, \underline{\underline{\Sigma}}).
$${#eq-vi_full}
